---
title: "Taxane metabolic pathway"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## ALPS results

This is meant to generate the results from ALPS; ideally would be able to run ALPS in one place, then this file can be rendered.

Still need code for:

-   Table 1/study population descriptives

-   Number of events in study population (?)

```{r}
#| label: setup
#| warning: false
#| message: false

library(here)
library(tidyverse)
# for ALPS:
#library(ape)
#library(phytools)
#library(geiger)
library(data.table)
#library(stringr)
# other 
library(survival)
library(ggplot2)
library(gtsummary)
library(gt)
# load ALPS functions
source(file = paste0("./alps2.R")) # alps
```

Summarize results

```{r}
#| label: ALPS-summarize-posterior
fstree <- fread(paste0("./output/logs-tree.txt"), sep = ";", header = FALSE)
fssampler <- fread(paste0("./output/logs-sampler.txt"), sep = "\t", header = TRUE)
fspsi <- fread(paste0("./output/logs-psi.txt"), sep = "\t", header = TRUE)
fsparameters <- fread(paste0("./output/logs-tree-parameters.txt"), sep = "\t", header = TRUE)
fsthetas <- fread(paste0("./output/logs-tree-thetas.txt"), sep = "\t", header = TRUE)

# check hyperparameter, looking for a spread/range 
table(fspsi$psi)

# completed iterations
niter <- nrow(fssampler)
niter

# acceptance rate for proposed changes
length(fssampler$accept[fssampler$accept == TRUE])/nrow(fssampler)

# number of internal nodes
table(fssampler$nnodes/nrow(fssampler))

# distance from prior forest
table(fssampler$distance)
```

#### Tree structures

```{r}
#| label: tree-structures-posterior

# link to sampler
fstree$V2 <- NULL
fstree$treeindex <- as.numeric(substr(fstree$V1,1,regexpr('\\(', fstree$V1)-1))
fstree$tree <- substr(fstree$V1,regexpr('\\(',fstree$V1), nchar(fstree$V1))
fssamplertree <- merge(fssampler,fstree,by.x="treeindex",by.y="treeindex",all.x=T,sort=F)

# tree posteriors
treepost <- aggregate(fssamplertree$tree, by=list(fssamplertree$tree), FUN=function(x){NROW(x)})
names(treepost) <- c("tree","count")
treepost$post <- treepost$count/nrow(fssampler)
treepost$perc <- treepost$post*100
# TODO confirm this is fine and not referring to prior-only run
treeprior <- mean(treepost$post)
treepost$BF <- round((treepost$post/(1-treepost$post))/(treeprior/(1-treeprior)),0)

# print(treepost)
# gt(treepost)
treepost %>%
  filter(count > 1000) %>%
  select(-perc) %>%
  arrange(desc(BF)) %>%
  gt() %>%
  tab_header(title = "ALPS Tree Posteriors and BF > 1")
```

The above is a subset because there were a LOT of tree posteriors. Including at the end of this document.

#### Posterior odds for individual variants

```{r}
#| label: posterior-odds
#| tbl-cap: "Marginal posteriors"
text <- paste(treepost$tree,";",sep = "")
temp <- read.tree(text = text)
getlabels <- function(tree){
  return(tree$tip.label)
}

tips <- lapply(temp, getlabels)
nt <- unique(unlist(tips))
main <- matrix(0, ncol = length(nt))
names(main) <- nt
for (i in 1:nrow(treepost)){
  main[tips[[i]]] <- main[tips[[i]]] + treepost$count[i]
}
mainpost <- data.frame(count = main, post = main/niter)

# Bayes factors (BF)
# TODO confirm that this should be created from this vs. from prior files 
mainprior <- mean(mainpost$post)
mainpost$bf <- round((mainpost$post/(1 - mainpost$post))/
                       (mainprior/(1 - mainprior)),0)

# Genetic factor marginal posteriors
# mainpost
# TODO add column names 
gt(mainpost,
   rownames_to_stub = TRUE)
```

```{r}
#| tbl-cap: "Trees with BF>=1"
#| label: post-cladograms

# parameters for any tree with BF >= 1
treepost$perc <- round(treepost$post*100)
s <- subset(treepost,BF>=1)
gt(s)

s2 <- subset(mainpost, bf >= 1)

# plot cladograms of selected tree structures
par(mfrow = c(2,2))

s3 <- data.frame(
  tree = character(),
  beta = numeric(),
  theta.1 = numeric(),
  theta.2 = numeric(),
  stringsAsFactors = FALSE
)

for (i in 1:nrow(s)) {
	print(s$tree[i])
	ti <- min(fstree[fstree$tree == s$tree[i]]$treeindex)
	tiparam <- subset(fsparameters, treeindex == ti)
	print(tiparam)
	tithetas <- subset(fsthetas, treeindex == ti)
	print(tithetas)
	t <- read.tree(text = paste0(s$tree[i], ";"))
	plot(t,root.edge = TRUE, main = paste0("Posterior: ", s$perc[i], "% BF:", s$BF[i]), type = "cladogram", sub = paste0("beta=", round(tiparam$beta, 2)))
	edgelabels(round(as.vector(t(subset(tithetas, select = c("theta.1", "theta.2")))), 2))
	par(ask=TRUE) 
	  s3 <- rbind(s3, data.frame(
    tree = s$tree[i],
    treeindex = tiparam$treeindex,
    beta = tiparam$beta,
    nnodes = tiparam$nnodes,
    negloglike = tiparam$negloglik,
    logpost = tiparam$logpost,
    distance = tiparam$distance,
    theta.1 = tithetas$theta.1,
    theta.2 = tithetas$theta.2
  ))
}

s3 %>%
  mutate(beta = round(beta, 2),
         theta.1  = round(theta.1, 2),
         theta.2  = round(theta.2, 2)) %>%
  gt()
```

Looks like we do get trees, though simple.

```{r}
#| label: concept-posterior

path <- read.csv("./taxane_summary.csv", header = TRUE)

path_names <- names(path)[4:8]
concept_names <- data.frame()
for (i in path_names){
  cur <- subset(path, !is.na(path[i]))
  
  concept.post <- 0
  for (j in 1:nrow(treepost)){
    test <- any(tips[[j]] %in% cur$datavar)
    if(test) {
      concept.post <- concept.post + treepost$count[j]
    }	
  }
  concept.post <- concept.post/niter
  print(paste0("concept:", i , " posterior:", concept.post))
  concept_names <- rbind(concept_names, data.frame(
    concept = i,
    posterior = concept.post
  ))
}
gt(concept_names)
```

Next steps:

To calculate Bayes factors for the pathway concepts:

-   generate concept probabilities from a prior-only ALPS run, and convert to prior odds (P/1-P)

-   calculate posterior odds from the posterior probabilities in the output above

-   divide the posterior odds by the prior odds to yield the Bayes factor

```{r}
#| label: prior-odds
#| eval: false
#| include: false

prior_fstree <- fread(paste0("./priorAK/prior-tree.txt"), sep = ";", header = FALSE)
prior_fssampler <- fread(paste0("./priorAK/prior-sampler.txt"), sep = "\t", header = TRUE)
prior_fsparameters <- fread(paste0("./priorAK/prior-tree-parameters.txt"), sep = "\t", header = TRUE)

prior_niter <- nrow(prior_fssampler)
prior_niter

prior_fstree$V2 <- NULL
prior_fstree$treeindex <- as.numeric(substr(prior_fstree$V1,1,regexpr('\\(', prior_fstree$V1)-1))
prior_fstree$tree <- substr(prior_fstree$V1,regexpr('\\(',prior_fstree$V1), nchar(prior_fstree$V1))
prior_fssamplertree <- merge(prior_fssampler, prior_fstree, by.x="treeindex", by.y="treeindex", all.x=T, sort=F)

prior_treepost <- aggregate(prior_fssamplertree$tree, by=list(prior_fssamplertree$tree), FUN=function(x){NROW(x)})
names(prior_treepost) <- c("tree","count")
prior_treepost$post <- prior_treepost$count/nrow(prior_fssampler)
prior_treepost$perc <- prior_treepost$post*100
prior_treeprior <- mean(prior_treepost$post)
prior_treepost$BF <- round((prior_treepost$post/(1-prior_treepost$post))/(prior_treeprior/(1-prior_treeprior)),0)

prior_niter <- nrow(prior_fssampler)
prior_niter
# adapting code from posterior odds for individual variants
prior_text <- paste(prior_treepost$tree,";",sep = "")
prior_temp <- read.tree(text = prior_text)
# getlabels created earlier 
prior_tips <- lapply(temp, getlabels)
prior_nt <- unique(unlist(prior_tips))
prior_main <- matrix(0, ncol = length(prior_nt))
names(prior_main) <- prior_nt
for (i in 1:nrow(prior_treepost)){
  main[prior_tips[[i]]] <- prior_main[prior_tips[[i]]] + prior_treepost$count[i]
}
# prior_mainpost <- data.frame(count = main, post = main/niter)

# Bayes factors (BF)
# TODO confirm that this should be created from this vs. from prior files 
# mainprior <- mean(mainpost$post)
# mainpost$bf <- round((mainpost$post/(1 - mainpost$post))/
#                       (mainprior/(1 - mainprior)),0)
```

I compared results with both me running 1 million with prior and dummy data (since in theory, observed data not used) but also with 100 iterations that Anders ran; both look very similar. I am wondering if I am using the wrong column in the prior tree post.

```{r}
#| label: concepts-all
path <- read.csv("./taxane_summary.csv", header = TRUE)

path_names <- names(path)[4:8]
concept_names <- data.frame()
for (i in path_names){
  cur <- subset(path, !is.na(path[i]))

  concept.prior <- 0
  for (j in 1:nrow(prior_treepost)){
    test <- any(prior_tips[[j]] %in% cur$datavar)
    if(test) {
      concept.prior <- concept.prior + prior_treepost$count[j]
    }   
  }
      concept.post <- 0
  for (j in 1:nrow(treepost)){
    test <- any(tips[[j]] %in% cur$datavar)
    if(test) {
      concept.post <- concept.post + treepost$count[j]
    }
  }
  concept.post <- concept.post/niter
  concept.prior <- concept.prior/prior_niter
  print(paste0("concept:", i , " posterior:", concept.post, "prior:", concept.prior))
  concept_names <- rbind(concept_names, data.frame(
    concept = i,
    posterior_prob = concept.post,
    prior_prob = concept.prior
  ))
}

# TODO why are my priors 0 and 1
concept_table <- concept_names %>%
  mutate(posterior_odds = posterior_prob/(1-posterior_prob),
         prior_odds = prior_prob/(1-prior_prob),
         BF = posterior_odds/prior_odds) 

concept_table %>%
  select(concept, posterior_prob, prior_prob, BF) %>%
  gt()
```

A filtered version of this table is presented above, this is the full table of posterior trees:

```{r}
#| label: all-post-trees
#| tbl-cap: "Tree posteriors and BFs (all)"
treepost %>%
  arrange(desc(BF)) %>%
  gt()
```
